---
title: "Using spike-and-slab joint graphical lasso"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using spike-and-slab joint graphical lasso}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The `spikeyglass` package implements the Bayesian Spike-and-Slab Joint
Graphical Lasso (SSJGL) from Li et al. (2019). This method estimates
multiple related precision matrices (inverse covariance matrices) across
K groups simultaneously, encouraging shared sparsity structure while
allowing group-specific differences.

Key features:

- **Adaptive penalization**: spike-and-slab priors automatically determine
  edge-specific penalty weights
- **Two penalty types**: fused (penalizes pairwise differences) and group
  (penalizes L2 norm across groups)
- **Dynamic posterior exploration**: a ladder of v0 values avoids expensive
  cross-validation
- **Missing data handling**: built-in imputation via conditional MVN

## Simulate data

We generate a small synthetic example with K=2 groups and p=15 variables
using a banded (AR-2) graph with a few differential edges.

```{r simulate}
library(spikeyglass)

sim <- simulate_ssjgl_data(
  K = 2, p = 15, n = 100,
  graph_type = "band",
  perturb_prob = 0.05,
  signal = 0.3,
  seed = 42
)

# True adjacency matrices
cat("Group 1 edges:", sum(sim$adj_list[[1]][upper.tri(sim$adj_list[[1]])]), "\n")
cat("Group 2 edges:", sum(sim$adj_list[[2]][upper.tri(sim$adj_list[[2]])]), "\n")

# Visualize true graph structures
par(mfrow = c(1, 2))
image(sim$adj_list[[1]], main = "True Graph: Group 1",
      col = c("white", "steelblue"), axes = FALSE)
image(sim$adj_list[[2]], main = "True Graph: Group 2",
      col = c("white", "steelblue"), axes = FALSE)
```

## Choosing hyperparameters

SSJGL has three penalty parameters (`lambda0`, `lambda1`, `lambda2`) and
a spike variance ladder (`v0s`). Understanding how to set these is key to
getting good results.

### Lambda parameters

The spike-and-slab E-step produces adaptive, edge-specific penalty weights
that multiply the base `lambda1` and `lambda2` values. This means the method
is **relatively insensitive to the absolute values** of lambda1 and lambda2
(Li et al., 2019) --- the adaptive weights compensate. What matters most is:

1. **Scale of lambda1**: Should match the scale of your data.
   - **Normalized data** (`normalize = TRUE`): use `lambda1 = 0.01` to `0.1`
   - **Raw data**: use `lambda1 = 0.5` to `1`

2. **Ratio lambda2/lambda1**: Controls the balance between within-group
   sparsity and cross-group similarity.
   - `lambda2 = 0`: No borrowing across groups (separate estimation)
   - `lambda2 = lambda1`: Equal weight on sparsity and similarity
   - `lambda2 > lambda1`: Encourage more similar graphs across groups

3. **lambda0**: Diagonal penalty. Use `lambda0 = 1` (the method is insensitive
   to this choice).

### The v0 ladder

The spike variance `v0` is the **primary model selection parameter**. The
effective spike penalty is `lambda1/v0`, so `v0` must be scaled relative to
`lambda1`. The `make_v0_ladder()` function handles this automatically:

```{r v0_ladder}
# Generate a v0 ladder appropriate for lambda1 = 0.5
lambda1 <- 0.5
v0s <- make_v0_ladder(lambda1, n_steps = 10, max_mult = 200)

cat("v0 range:", round(range(v0s), 5), "\n")
cat("Effective penalty range (lambda1/v0):",
    round(range(lambda1 / v0s), 1), "\n")
```

The `max_mult` parameter controls how sparse the final model gets:

- `max_mult = 50`: moderate sparsity
- `max_mult = 200`: strong sparsity (default)
- `max_mult = 1000`: very aggressive sparsity

## Fit the model

We use `normalize = TRUE` (data is not pre-standardized), `lambda1 = 0.5`,
`lambda2 = 0.5`, and a v0 ladder scaled to lambda1:

```{r fit, results='hide'}
v0s <- make_v0_ladder(lambda1 = 0.5, n_steps = 10, max_mult = 200)

fit <- ssjgl(
  Y = sim$data_list,
  penalty = "fused",
  lambda0 = 1,
  lambda1 = 0.5,
  lambda2 = 0.5,
  v1 = 1,
  v0s = v0s,
  doubly = TRUE,
  a = 1, b = 15,
  maxitr.em = 100,
  tol.em = 1e-4,
  normalize = TRUE,
  impute = FALSE
)
```

```{r print}
print(fit)
```

## Assessing stability

Use `plot_stability()` to see how the graph structure evolves across the
v0 ladder. Look for the point where edge counts and inclusion probabilities
**stop changing** --- this indicates the solution has converged.

```{r stability, fig.height=4}
plot_stability(fit, v0s)
```

When the "Edges changed" panel drops to zero (or near zero), the solution
has stabilized. If it hasn't stabilized by the last v0 step, try increasing
`max_mult` in `make_v0_ladder()` to push the ladder further.

## Visualize the solution path

The solution path shows how individual precision matrix entries evolve.
As v0 decreases, edges with low inclusion probability are shrunk to zero:

```{r path, fig.height=4}
plot_path(v0s, fit, thres = 0.5, normalize = FALSE,
          xlab = expression(v[0]), ylab = "Precision entries",
          main = c("Group 1", "Group 2"), par = c(1, 2))
```

## Extract results

```{r extract}
# Precision matrices at the final (most sparse) v0 step
theta_hat <- coef(fit)

# Partial correlations
pcor_hat <- fitted(fit)

# Edge inclusion probabilities
probs <- extract_probabilities(fit)

# Binary adjacency from thresholding inclusion probabilities
adj_hat <- extract_adjacency(fit, threshold = 0.5)

cat("Estimated edges (Group 1):",
    sum(adj_hat[[1]][upper.tri(adj_hat[[1]])]), "\n")
cat("Estimated edges (Group 2):",
    sum(adj_hat[[2]][upper.tri(adj_hat[[2]])]), "\n")
```

## Evaluate performance

Compare the estimated graphs to the truth using TPR, FPR, and AUC:

```{r metrics}
metrics <- compute_metrics(
  fit = fit,
  true_adj = sim$adj_list,
  true_omega = sim$Omega_list,
  threshold = 0.5
)

cat("=== Overall Performance ===\n")
cat(sprintf("Mean TPR: %.3f\n", metrics$overall$mean_TPR))
cat(sprintf("Mean FPR: %.3f\n", metrics$overall$mean_FPR))
cat(sprintf("Mean AUC: %.3f\n", metrics$overall$mean_AUC))

for (k in 1:2) {
  cat(sprintf("\n--- Group %d ---\n", k))
  g <- metrics$per_group[[k]]
  cat(sprintf("  TP=%d, FP=%d, TN=%d, FN=%d\n", g$TP, g$FP, g$TN, g$FN))
  cat(sprintf("  TPR=%.3f, FPR=%.3f, F1=%.3f\n", g$TPR, g$FPR, g$F1))
  cat(sprintf("  Frobenius norm: %.3f\n", g$frobenius_norm))
}
```

### ROC curve

```{r roc, fig.height=5, fig.width=5}
plot_roc(metrics$roc[[1]], main = "ROC Curve (Group 1)")
```

## Select v0 via cross-validation

For real data where the truth is unknown, use cross-validation to select
the best v0. This evaluates each v0 independently using held-out Gaussian
negative log-likelihood:

```{r cv, eval=FALSE}
v0s_cv <- make_v0_ladder(lambda1 = 0.5, n_steps = 8)

cv_res <- SSJGL_select_v0_cv(
  Y = sim$data_list,
  v0s = v0s_cv,
  folds = 3,
  penalty = "fused",
  lambda0 = 1, lambda1 = 0.5, lambda2 = 0.5,
  maxitr.em = 50, maxitr.jgl = 50,
  normalize = TRUE
)
cat("Best v0:", cv_res$v0_best, "\n")
```

## Bootstrap confidence intervals

The `SSJGL_final_with_pcor_CI` function computes bootstrap confidence
intervals for partial correlations at a given v0:

```{r bootstrap, eval=FALSE}
boot_res <- SSJGL_final_with_pcor_CI(
  Y = sim$data_list,
  v0_best = 0.005,
  B = 50,
  ci_level = 0.95,
  penalty = "fused",
  lambda0 = 1, lambda1 = 0.5, lambda2 = 0.5,
  normalize = TRUE
)

# Partial correlation point estimate for group 1
pcor_g1 <- boot_res$pcor_hat[[1]]

# 95% CI bounds
lo <- boot_res$CI_lower[[1]]
hi <- boot_res$CI_upper[[1]]

# Edges where CI excludes zero
significant_edges <- (lo > 0 | hi < 0)
diag(significant_edges) <- FALSE
cat("Significant edges (Group 1):",
    sum(significant_edges[upper.tri(significant_edges)]), "\n")
```

## Tuning summary

| Parameter | Role | Guidance |
|-----------|------|----------|
| `lambda0` | Diagonal penalty | Use 1 (insensitive) |
| `lambda1` | Edge sparsity | 0.01--0.1 (normalized), 0.5--1 (raw) |
| `lambda2` | Cross-group similarity | Same scale as lambda1; ratio controls borrowing |
| `v0s` | Model selection ladder | Use `make_v0_ladder(lambda1)` |
| `v1` | Slab variance | Leave at 1 |
| `a, b` | Beta prior for pi | `a = 1, b = p` for sparse prior |
| `doubly` | Doubly spike-and-slab | TRUE recommended for multi-group |
| `normalize` | Mean-center variables | TRUE unless data is pre-standardized |

**Recommended workflow:**

1. Set `lambda0 = 1`, choose `lambda1` based on data scale
2. Set `lambda2 = lambda1` as starting point (adjust ratio for more/less
   cross-group similarity)
3. Use `v0s = make_v0_ladder(lambda1)` for the v0 ladder
4. Fit with `ssjgl()`, then check `plot_stability()` for convergence
5. If the path hasn't stabilized, increase `max_mult`
6. For real data: use `SSJGL_select_v0_cv()` or `plot_path()` to pick v0

## References

Li, Z. R., McCormick, T. H., & Clark, S. J. (2019). Bayesian Joint
Spike-and-Slab Graphical Lasso. *Proceedings of the 36th International
Conference on Machine Learning (ICML)*.
